---
title: "Introduction to vivaglint"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to vivaglint}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

**vivaglint** provides a comprehensive toolkit for analyzing Microsoft Viva Glint survey data exports. The package handles data import, validation, statistical analysis, and reporting with specialized functions for common survey analysis tasks.

## Key Features

- **Data Import & Validation**: Automatic validation of Glint export structure
- **Summary Statistics**: Mean, SD, response rates, favorability percentages
- **Multi-cycle Analysis**: Track changes over time
- **Organizational Analysis**: Roll up results by manager hierarchy
- **Advanced Analytics**: Correlation analysis, factor analysis, attrition prediction
- **Demographic Segmentation**: Analyze results by any employee attributes

## Getting Started

### Installation

```{r eval=FALSE}
# Install from GitHub
devtools::install_github("microsoft/vivaglint")

# Load the package
library(vivaglint)
```

### Basic Workflow

#### 1. Import Survey Data

The `read_glint_survey()` function imports and validates your Glint CSV export:

```{r eval=FALSE}
survey <- read_glint_survey("survey_export.csv")
```

This returns a `glint_survey` object containing:
- **data**: The survey responses
- **metadata**: Question names, respondent counts, etc.

#### 2. Summarize Questions

Calculate comprehensive metrics for all questions:

```{r eval=FALSE}
summary <- summarize_survey(survey, scale_points = 5)
```

This returns a tibble with one row per question containing:
- `mean` - Average response score
- `sd` - Standard deviation
- `n_responses` - Number of responses
- `response_rate` - Response rate (0-1)
- `pct_favorable` - Percentage of favorable responses
- `pct_neutral` - Percentage of neutral responses
- `pct_unfavorable` - Percentage of unfavorable responses

**Favorability Classification**: The package uses Glint's standard favorability mappings for 2-11 point scales. For example, on a 5-point scale, responses of 4 or 5 are favorable, 3 is neutral, and 1-2 are unfavorable.

#### 3. Analyze Specific Questions

Focus on particular questions of interest:

```{r eval=FALSE}
engagement_qs <- c("My work is meaningful", "I feel valued at work")
result <- summarize_survey(survey,
                           scale_points = 5,
                           questions = engagement_qs)
```

#### 4. Get Response Distributions

See the detailed breakdown of response values:

```{r eval=FALSE}
distributions <- get_response_dist(survey)
```

This expands value counts into separate columns like `count_1`, `count_2`, `pct_1`, `pct_2`, etc.

## Multi-Cycle Comparisons

Track how scores change across survey cycles:

```{r eval=FALSE}
# Load multiple survey cycles
survey_q1 <- read_glint_survey("2024_Q1.csv")
survey_q2 <- read_glint_survey("2024_Q2.csv")
survey_q3 <- read_glint_survey("2024_Q3.csv")

# Compare across cycles
comparison <- compare_cycles(survey_q1, survey_q2, survey_q3,
                             scale_points = 5,
                             cycle_names = c("Q1 2024", "Q2 2024", "Q3 2024"))
```

The output includes:
- All metrics from `summarize_survey()` for each cycle
- `change_from_previous` - Change in mean score
- `pct_change_from_previous` - Percentage change

## Organizational Analysis

### Aggregate by Manager

Roll up survey results to manager level:

```{r eval=FALSE}
# Direct reports only
manager_summary <- aggregate_by_manager(survey, scale_points = 5)

# Include full organizational tree
manager_full <- aggregate_by_manager(survey,
                                     scale_points = 5,
                                     full_tree = TRUE)
```

Returns metrics for each manager's team including:
- `manager_id` and `manager_name`
- `team_size` - Number of team members
- All standard survey metrics

## Advanced Analytics

### Correlation Analysis

Examine relationships between questions:

```{r eval=FALSE}
# Calculate correlations (default: Spearman)
correlations <- get_correlations(survey, method = "spearman")

# Get correlation matrix instead of long format
cor_matrix <- get_correlations(survey, format = "matrix")
```

Supported methods: Spearman, Pearson, Kendall

### Factor Analysis

Identify latent constructs in your survey:

```{r eval=FALSE}
# Requires psych package
factors <- extract_survey_factors(survey,
                                  n_factors = 3,
                                  rotation = "oblimin")

# View loadings
print(factors$loadings_clean)

# View variance explained
print(factors$variance_explained)
```

### Attrition Analysis

Link survey responses to employee turnover:

```{r eval=FALSE}
attrition <- analyze_attrition(
  survey,
  attrition_file = "employee_attributes.csv",
  emp_id_col = "EMP ID",
  term_date_col = "Termination Date",
  scale_points = 5,
  time_periods = c(30, 90, 180)  # days
)
```

Returns attrition rates for favorable vs unfavorable responders with risk ratios.

### Demographic Analysis

Segment results by employee attributes:

```{r eval=FALSE}
demo_results <- analyze_by_attributes(
  survey,
  attribute_file = "employee_attributes.csv",
  scale_points = 5,
  attribute_cols = c("Department", "Gender", "Tenure Group"),
  min_group_size = 10  # Privacy threshold
)
```

Analyzes survey results for all combinations of the specified attributes, filtering out groups below the minimum size threshold.

## Data Reshaping

Convert to long format for custom analysis:

```{r eval=FALSE}
# All responses in long format
long_data <- pivot_long(survey, data_type = "all")

# Comments only
comments_data <- pivot_long(survey, data_type = "comments")

# Both as separate tibbles
both <- pivot_long(survey, data_type = "both")
long_responses <- both$responses
long_comments <- both$comments
```

## Best Practices

### Privacy Protection

- Use `min_group_size` parameters to prevent identification of small groups
- Default minimum is 5 employees per group
- Consider higher thresholds (10-15) for sensitive analyses

### Scale Points

Always specify the correct `scale_points` parameter:
- Common scales: 2, 3, 5, 7, 10, 11
- Determines favorability classification
- Critical for accurate reporting

### Data Quality

The package validates:
- Required column structure
- Date formats
- Question completeness
- Provides informative error messages for data issues

## Additional Resources

- **Function Documentation**: `?read_glint_survey`, `?summarize_survey`, etc.
- **GitHub**: https://github.com/microsoft/vivaglint
- **Issues**: https://github.com/microsoft/vivaglint/issues

## Data Privacy

This package processes all data locally within your R environment. No data is transmitted to external services, including Microsoft. Always follow your organization's data handling policies when working with employee survey data.
